\chapter{``A'' Standard Extension for Atomic Instructions, Version 2.0}
\label{atomics}

The standard atomic-instruction extension, named ``A'',
contains instructions that atomically
read-modify-write memory to support synchronization between multiple
RISC-V harts running in the same memory space.  The two forms of
atomic instruction provided are load-reserved/store-conditional
instructions and atomic fetch-and-op memory instructions.  Both types
of atomic instruction support various memory consistency orderings
including unordered, acquire, release, and sequentially consistent
semantics.  These instructions allow RISC-V to support the RCsc memory
consistency model~\cite{Gharachorloo90memoryconsistency}.

\begin{commentary}
After much debate, the language community and architecture community
appear to have finally settled on release consistency as the standard
memory consistency model and so the RISC-V atomic support is built
around this model.
\end{commentary}

\section{Specifying Ordering of Atomic Instructions}

The base RISC-V ISA has a relaxed memory model, with the FENCE
instruction used to impose additional ordering constraints.  The
address space is divided by the execution environment into memory and
I/O domains, and the FENCE instruction provides options to order
accesses to one or both of these two address domains.

To provide more efficient support for release
consistency~\cite{Gharachorloo90memoryconsistency}, each atomic
instruction has two bits, {\em aq} and {\em rl}, used to specify
additional memory ordering constraints as viewed by other RISC-V
harts.  The bits order accesses to one of the two address domains,
memory or I/O, depending on which address domain the atomic
instruction is accessing.  No ordering constraint is implied to
accesses to the other domain, and a FENCE instruction should be used
to order across both domains.

If both bits are clear, no additional ordering constraints are imposed
on the atomic memory operation.  If only the {\em aq} bit is set, the
atomic memory operation is treated as an {\em acquire} access, i.e.,
no following memory operations on this RISC-V hart can be observed
to take place before the acquire memory operation.  If only the {\em
  rl} bit is set, the atomic memory operation is treated as a {\em
  release} access, i.e., the release memory operation cannot be
observed to take place before any earlier memory operations on this
RISC-V hart.  If both the {\em aq} and {\em rl} bits are set, the
atomic memory operation is {\em sequentially consistent} and cannot be
observed to happen before any earlier memory operations or after any
later memory operations in the same RISC-V hart and to the same
address domain.

\section{Load-Reserved/Store-Conditional Instructions}
\label{sec:lrsc}

\vspace{-0.2in}
\begin{center}
\begin{tabular}{R@{}W@{}W@{}R@{}R@{}F@{}R@{}O}
\\
\instbitrange{31}{27} &
\instbit{26} &
\instbit{25} &
\instbitrange{24}{20} &
\instbitrange{19}{15} &
\instbitrange{14}{12} &
\instbitrange{11}{7} &
\instbitrange{6}{0} \\
\hline
\multicolumn{1}{|c|}{funct5} &
\multicolumn{1}{c|}{aq} &
\multicolumn{1}{c|}{rl} &
\multicolumn{1}{c|}{rs2} &
\multicolumn{1}{c|}{rs1} &
\multicolumn{1}{c|}{funct3} &
\multicolumn{1}{c|}{rd} &
\multicolumn{1}{c|}{opcode} \\
\hline
5 & 1 & 1 & 5 & 5 & 3 & 5 & 7 \\
LR.W/D & \multicolumn{2}{c}{ordering} & 0   & addr & width & dest & AMO    \\
SC.W/D & \multicolumn{2}{c}{ordering} & src & addr & width & dest & AMO  \\
\end{tabular}
\end{center}

Complex atomic memory operations on a single memory word or doubleword are performed
with the load-reserved (LR) and store-conditional (SC) instructions.
LR.W loads a word from the address in {\em rs1}, places the
sign-extended value in {\em rd}, and registers a reservation on the
memory address and a range of bytes including at least all bytes of
the addressed word.  SC.W writes a word in {\em rs2} to the address in
{\em rs1}, provided a valid reservation still exists on that address.
SC.W writes zero to {\em rd} on success or a nonzero code on failure.
LR.D and SC.D act analogously on doublewords and are only available on RV64.
For RV64, LR.W and SC.W sign-extend the value placed in {\em rd}.

\begin{commentary}
Both compare-and-swap (CAS) and LR/SC can be used to build lock-free
data structures.  After extensive discussion, we opted for LR/SC for
several reasons: 1) CAS suffers from the ABA problem, which LR/SC
avoids because it monitors all accesses to the address rather than
only checking for changes in the data value; 2) CAS would also require
a new integer instruction format to support three source operands
(address, compare value, swap value) as well as a different memory
system message format, which would complicate microarchitectures; 3)
Furthermore, to avoid the ABA problem, other systems provide a
double-wide CAS (DW-CAS) to allow a counter to be tested and
incremented along with a data word. This requires reading five
registers and writing two in one instruction, and also a new larger
memory system message type, further complicating implementations; 4)
LR/SC provides a more efficient implementation of many primitives as
it only requires one load as opposed to two with CAS (one load before
the CAS instruction to obtain a value for speculative computation,
then a second load as part of the CAS instruction to check if value is
unchanged before updating).

The main disadvantage of LR/SC over CAS is livelock, which we avoid,
under certain circumstances,
with an architected guarantee of eventual forward progress as
described below.  Another concern is whether the influence of the
current x86 architecture, with its DW-CAS, will complicate porting of
synchronization libraries and other software that assumes DW-CAS is
the basic machine primitive.  A possible mitigating factor is the
recent addition of transactional memory instructions to x86, which
might cause a move away from DW-CAS.
\end{commentary}

The failure code with value 1 is reserved to encode an unspecified
failure.  Other failure codes are reserved at this time, and portable
software should only assume the failure code will be non-zero.

\begin{commentary}
We reserve a failure code of 1 to mean ``unspecified'' so that simple
implementations may return this value using the existing mux required
for the SLT/SLTU instructions.  More specific failure codes might be
defined in future versions or extensions to the ISA.
\end{commentary}

For LR and SC, the A extension requires that the address held in {\em
  rs1} be naturally aligned to the size of the operand (i.e.,
eight-byte aligned for 64-bit words and four-byte aligned for 32-bit
words).  If the address is not naturally aligned, an address-misaligned
exception or an access-fault exception will be generated.  The access-fault
exception can be generated for a memory access that would otherwise be
able to complete except for the misalignment, if the misaligned access
should not be emulated.

An implementation can reserve an arbitrarily large subset of the
address space on each LR, provided the memory range includes all bytes
of the addressed data word or doubleword.
An SC can only pair with the most recent LR in program order.  An SC may
succeed if no store from another hart, nor a write from some other device, to
the address range reserved by the LR can be observed to have occurred between
the LR and the SC, and if there is no other SC between the LR and itself in
program order.
Note this LR might have had a different address
argument and data size, but reserved the SC's address as part of the memory subset.
Following this model, in systems with memory translation, an SC is
allowed to succeed if the earlier LR reserved the same location using
an alias with a different virtual address, but is also allowed to fail
if the virtual address is different.
The SC must fail if the address is not within the memory subset reserved
by the most recent LR in program order.
The SC must fail if a store from another hart, or a write from some other
device, to the address range reserved by the LR can be observed to occur
between the LR and the SC.
An SC must fail if there is
another SC (to any address) between the LR and the SC in program
order.  The precise statement of the atomicity requirements for
successful LR/SC sequences is defined by the Atomicity Axiom in
Section~\ref{sec:rvwmo}.

\begin{commentary}
The platform should provide a means to determine the size and shape of the
memory range reserved on an LR.

A platform specification may constrain the size and shape of the memory range
reserved on an LR.  For example, the Unix platform is expected to require the
range be contiguous and no greater than the virtual memory page size.
\end{commentary}

\begin{commentary}
A store-conditional instruction to a scratch word of memory should be used
during a preemptive context switch to forcibly yield any existing load
reservation.
\end{commentary}

LR/SC can be used to construct lock-free data structures.  An example
using LR/SC to implement a compare-and-swap function is shown in
Figure~\ref{cas}.  If inlined, compare-and-swap functionality need
only take four instructions.

\begin{figure}[h!]
\begin{center}
\begin{verbatim}
        # a0 holds address of memory location 
        # a1 holds expected value
        # a2 holds desired value
        # a0 holds return value, 0 if successful, !0 otherwise
    cas:
        lr.w t0, (a0)        # Load original value.
        bne t0, a1, fail     # Doesn't match, so fail.
        sc.w t0, a2, (a0)    # Try to update.
        bnez t0, cas         # Retry if store-conditional failed.
        li a0, 0             # Set return to success.
        jr ra                # Return.
    fail:
        li a0, 1             # Set return to failure.
        jr ra                # Return.
\end{verbatim}
\end{center}
\caption{Sample code for compare-and-swap function using LR/SC.}
\label{cas}
\end{figure}

An SC instruction can never be observed by another RISC-V hart
before the immediately preceding LR.  Due to the atomic nature of the
LR/SC sequence, no memory operations from another hart can be observed
to have occurred between the LR and a successful SC.  The LR/SC
sequence can be given acquire semantics by setting the {\em aq} bit on
the LR instruction.  The LR/SC sequence can be given release semantics
by setting the {\em rl} bit on the SC instruction.  Setting the {\em
  aq} bit on the LR instruction, and setting both the {\em aq} and the {\em
  rl} bit on the SC instruction makes the LR/SC sequence sequentially
consistent, meaning that it cannot be reordered with earlier or
later memory operations from the same hart.

If neither bit is set on both LR and SC, the LR/SC sequence can be
observed to occur before or after surrounding memory operations from
the same RISC-V hart.  This can be appropriate when the LR/SC
sequence is used to implement a parallel reduction operation.

Software should not set the {\em rl} bit on an LR instruction unless the {\em
aq} bit is also set, nor should software set the {\em aq} bit on an SC
instruction unless the {\em rl} bit is also set.  LR.{\em rl} and SC.{\em aq}
instructions are not guaranteed to provide any stronger ordering than those
with both bits clear, but may result in lower performance.

\begin{commentary}
In general, a multi-word atomic primitive is desirable but there is
still considerable debate about what form this should take, and
guaranteeing forward progress adds complexity to a system.  Our
current thoughts are to include a small limited-capacity transactional
memory buffer along the lines of the original transactional memory
proposals as an optional standard extension ``T''.
\end{commentary}

\section{Eventual Success of Store-Conditional Instructions}
\label{sec:lrscseq}

The standard A extension defines {\em constrained LR/SC loops}, which have
the following properties:
\vspace{-0.2in}
\begin{itemize}
\parskip 0pt
\itemsep 1pt
\item The loop comprises only an LR/SC sequence and code to retry the sequence
  in the case of failure, and must comprise at most 16 instructions placed
  sequentially in memory.
\item An LR/SC sequence begins with an LR instruction and ends with an SC
  instruction.  The dynamic code executed between the LR and SC instructions
  can only contain instructions from the base ``I'' instruction set, excluding
  loads, stores, backward jumps or taken backward branches, JALR, FENCE, and
  SYSTEM instructions.  If the ``C'' extension is supported, then compressed
  forms of the aforementioned ``I'' instructions are also permitted.
\item The code to retry a failing LR/SC sequence can contain backwards jumps
  and/or branches to repeat the LR/SC sequence, but otherwise has the same
  constraint as the code between the LR and SC.
\item The LR address must lie either within a main memory region or within some
  other memory region specified by the execution environment.
\item The SC must be to the same virtual address and of the same data size as
  the latest LR executed by the same hart.
\end{itemize}

LR/SC sequences that do not lie within constrained LR/SC loops are {\em
unconstrained}.  Unconstrained LR/SC sequences might succeed on some attempts
on some implementations, but might never succeed on other implementations.

\begin{commentary}
We restricted the length of LR/SC loops to fit within 64 contiguous
instruction bytes in the base ISA to avoid undue restrictions on instruction
cache and TLB size and associativity.
Similarly, we disallowed other loads and stores within the loops to avoid
restrictions on data-cache associativity in simple implementations that track
the reservation within the cache.
The restrictions on branches and jumps limit the time that
can be spent in the sequence.  Floating-point operations and integer
multiply/divide were disallowed to simplify the operating system's emulation
of these instructions on implementations lacking appropriate hardware support.

Software is not forbidden from using unconstrained LR/SC sequences, but
portable software must detect the case that the sequence repeatedly fails,
then fall back to an alternate code sequence that does not rely on an
unconstrained LR/SC sequence.  Implementations are permitted to
unconditionally fail any unconstrained LR/SC sequence.
\end{commentary}

If a hart {\em H} enters a constrained LR/SC loop, the execution environment
must guarantee that one of the following events eventually occurs:
\vspace{-0.2in}
\begin{itemize}
\parskip 0pt
\itemsep 1pt
\item {\em H} or some other hart executes a successful SC to the subset of
  memory reserved by the LR instruction in {\em H}'s constrained LR/SC loops.
\item Some other hart executes an unconditional store or AMO instruction to
  the subset of memory reserved by the LR instruction in {\em H}'s constrained
  LR/SC loop, or some other device in the system writes to that subset of
  memory.
\item {\em H} executes a branch or jump that exits the constrained LR/SC loop.
\item {\em H} executes an instruction that raises a synchronous exception.
\item {\em H} takes an interrupt.
\end{itemize}

\begin{commentary}
As a consequence of this requirement, if some harts in an execution
environment are executing constrained LR/SC loops, and no other harts or
devices in the execution environment execute an unconditional store or AMO to
that granule, then at least one hart will eventually exit its constrained
LR/SC loop.
By contrast, if other harts or devices continue to write to that granule,
it is not guaranteed that any hart will exit its LR/SC loop.

Loads and load-reserved instructions do not by themselves impede the progress
of other harts' LR/SC sequences.
We note this constraint implies that multithreaded cores require a mechanism
to prevent other threads' cache contention from precluding LR/SC progress.

These definitions admit the possibility that SC instructions may spuriously
fail for for implementation reasons, provided progress is eventually made.
\end{commentary}

\begin{commentary}
One advantage of CAS is that it guarantees that some hart eventually
makes progress, whereas an LR/SC atomic sequence could livelock
indefinitely on some systems.  To avoid this concern, we added an
architectural guarantee of livelock freedom for certain LR/SC sequences.

Earlier versions of this specification imposed a stronger starvation-freedom
guarantee.  However, the weaker livelock-freedom guarantee is sufficient to
implement the C11 and C++11 languages, and is substantially easier to provide
in some microarchitectural styles.
\end{commentary}

\section{Atomic Memory Operations}
\label{sec:amo}

\vspace{-0.2in}
\begin{center}
\begin{tabular}{O@{}W@{}W@{}R@{}R@{}F@{}R@{}R}
\\
\instbitrange{31}{27} &
\instbit{26} &
\instbit{25} &
\instbitrange{24}{20} &
\instbitrange{19}{15} &
\instbitrange{14}{12} &
\instbitrange{11}{7} &
\instbitrange{6}{0} \\
\hline
\multicolumn{1}{|c|}{funct5} &
\multicolumn{1}{c|}{aq} &
\multicolumn{1}{c|}{rl} &
\multicolumn{1}{c|}{rs2} &
\multicolumn{1}{c|}{rs1} &
\multicolumn{1}{c|}{funct3} &
\multicolumn{1}{c|}{rd} &
\multicolumn{1}{c|}{opcode} \\
\hline
5 & 1 & 1 & 5 & 5 & 3 & 5 & 7 \\
AMOSWAP.W/D & \multicolumn{2}{c}{ordering} & src & addr & width & dest & AMO  \\
AMOADD.W/D & \multicolumn{2}{c}{ordering} & src & addr & width & dest & AMO  \\
AMOAND.W/D & \multicolumn{2}{c}{ordering} & src & addr & width & dest & AMO  \\
AMOOR.W/D & \multicolumn{2}{c}{ordering} & src & addr & width & dest & AMO  \\
AMOXOR.W/D & \multicolumn{2}{c}{ordering} & src & addr & width & dest & AMO  \\
AMOMAX[U].W/D & \multicolumn{2}{c}{ordering} & src & addr & width & dest & AMO  \\
AMOMIN[U].W/D & \multicolumn{2}{c}{ordering} & src & addr & width & dest & AMO  \\
\end{tabular}
\end{center}

\vspace{-0.1in} The atomic memory operation (AMO) instructions perform
read-modify-write operations for multiprocessor synchronization and
are encoded with an R-type instruction format.  These AMO instructions
atomically load a data value from the address in {\em rs1}, place the
value into register {\em rd}, apply a binary operator to the loaded
value and the original value in {\em rs2}, then store the result back
to the address in {\em rs1}. AMOs can either operate on 64-bit (RV64
only) or 32-bit words in memory.  For RV64, 32-bit AMOs always
sign-extend the value placed in {\em rd}.

For AMOs, the A extension requires that the address held in {\em rs1}
be naturally aligned to the size of the operand (i.e., eight-byte
aligned for 64-bit words and four-byte aligned for 32-bit words).  If
the address is not naturally aligned, an address-misaligned exception
or an access-fault exception will be generated.  The access-fault exception can be
generated for a memory access that would otherwise be able to complete
except for the misalignment, if the misaligned access should not be
emulated.  The ``Zam'' extension, described in Chapter~\ref{sec:zam},
relaxes this requirement and specifies the semantics of misaligned
AMOs.

The operations supported are swap, integer add, bitwise AND, bitwise
OR, bitwise XOR, and signed and unsigned integer maximum and minimum.
Without ordering constraints, these AMOs can be used to implement
parallel reduction operations, where typically the return value would
be discarded by writing to {\tt x0}.

\begin{commentary}
We provided fetch-and-op style atomic primitives as they scale to
highly parallel systems better than LR/SC or CAS.  A simple
microarchitecture can implement AMOs using the LR/SC primitives.  More
complex implementations might also implement AMOs at memory
controllers, and can optimize away fetching the original value when
the destination is {\tt x0}.

The set of AMOs was chosen to support the C11/C++11 atomic memory
operations efficiently, and also to support parallel reductions in
memory.  Another use of AMOs is to provide atomic updates to
memory-mapped device registers (e.g., setting, clearing, or toggling
bits) in the I/O space.
\end{commentary}

To help implement multiprocessor synchronization, the AMOs optionally
provide release consistency semantics.  If the {\em aq} bit is set,
then no later memory operations in this RISC-V hart can be observed
to take place before the AMO.
Conversely, if the {\em rl} bit is set, then other
RISC-V harts will not observe the AMO before memory accesses
preceding the AMO in this RISC-V hart.  Setting both the {\em aq} and the {\em
rl} bit on an AMO makes the sequence sequentially consistent, meaning that
it cannot be reordered with earlier or later memory operations from the same
hart.

\begin{commentary}
The AMOs were designed to implement the C11 and C++11 memory models
efficiently.  Although the FENCE R, RW instruction suffices to
implement the {\em acquire} operation and FENCE RW, W suffices to
implement {\em release}, both imply additional unnecessary ordering as
compared to AMOs with the corresponding {\em aq} or {\em rl} bit set.
\end{commentary}

An example code sequence for a critical section guarded by a
test-and-test-and-set spinlock is shown in Figure~\ref{critical}.  Note the
first AMO is marked {\em aq} to order the lock acquisition before the
critical section, and the second AMO is marked {\em rl} to order
the critical section before the lock relinquishment.

\begin{figure}[h!]
\begin{center}
\begin{verbatim}
        li           t0, 1        # Initialize swap value.
    again:
        lw           t1, (a0)     # Check if lock is held.
        bnez         t1, again    # Retry if held.
        amoswap.w.aq t1, t0, (a0) # Attempt to acquire lock.
        bnez         t1, again    # Retry if held.
        # ...
        # Critical section.
        # ...
        amoswap.w.rl x0, x0, (a0) # Release lock by storing 0.
\end{verbatim}
\end{center}
\caption{Sample code for mutual exclusion.  {\tt a0} contains the address of the lock.}
\label{critical}
\end{figure}

\begin{commentary}
We recommend the use of the AMO Swap idiom shown above for both lock
acquire and release to simplify the implementation of speculative lock
elision~\cite{Rajwar:2001:SLE}.
\end{commentary}

The instructions in the ``A'' extension can also be used to provide
sequentially consistent loads and stores.  A sequentially consistent load can
be implemented as an LR with both {\em aq} and {\em rl} set. A sequentially
consistent store can be implemented as an AMOSWAP that writes the old value to
x0 and has both {\em aq} and {\em rl} set.
