\chapter{``N'' Standard Extension for User-Level Interrupts, Version 1.1}
\label{chap:n}

\begin{commentary}
  This is a placeholder for a more complete writeup of the N
  extension, and to form a basis for discussion.

  An ongoing topic of discussion is whether, for systems with only M and
  U privilege modes, the N extension should be supplanted by S-mode without
  virtual memory (i.e., with {\tt satp} hardwired to zero).
  This approach would have similar hardware cost and would simplify the
  architecture.
\end{commentary}

This chapter presents a proposal for adding RISC-V user-level
interrupt and exception handling.  When the N extension is present,
and the outer execution environment has delegated designated
interrupts and exceptions to user-level, then hardware can transfer
control directly to a user-level trap handler without invoking the
outer execution environment.

\begin{commentary}
User-level interrupts are primarily intended to support secure
embedded systems with only M-mode and U-mode present, but can also be
supported in systems running Unix-like operating systems to support
user-level trap handling.

When used in an Unix environment, the user-level interrupts would
likely not replace conventional signal handling, but could be used as
a building block for further extensions that generate user-level
events such as garbage collection barriers, integer overflow,
floating-point traps.
\end{commentary}

\section{Additional CSRs}

New user-visible CSRs added to support the N extension.
Their encodings are listed in Table~\ref{ucsrnames} in
Chapter~\ref{chap:priv-csrs}.

\subsection{User Status Register ({\tt ustatus})}

The {\tt ustatus} register is a UXLEN-bit read/write register
formatted as shown in Figure~\ref{ustatusreg}.  The {\tt ustatus}
register keeps track of and controls the hart's current operating
state.

\begin{figure*}[h!]
\begin{center}
\setlength{\tabcolsep}{4pt}
\begin{tabular}{KccFc}
\\
\instbitrange{UXLEN}{5} &
\instbit{4} &
\instbitrange{3}{1} &
\instbit{0} \\
\hline
\multicolumn{1}{|c|}{\wpri} &
\multicolumn{1}{c|}{UPIE} &
\multicolumn{1}{c|}{\wpri} &
\multicolumn{1}{c|}{UIE} \\
\hline
UXLEN-5 & 1 & 3 & 1 \\
\end{tabular}
\end{center}
\vspace{-0.1in}
\caption{User-mode status register ({\tt ustatus}).}
\label{ustatusreg}
\end{figure*}

The user interrupt-enable bit UIE disables user-level interrupts when
clear. The value of UIE is copied into UPIE when a user-level trap is
taken, and the value of UIE is set to zero to provide atomicity for
the user-level trap handler.

The UIE and UPIE bits are mirrored in the {\tt mstatus} and {\tt sstatus}
registers in the same bit positions.

\begin{commentary}
  There is no UPP bit to hold the previous privilege mode as it can
  only be user mode.
\end{commentary}

The URET instructions is used to return from traps in U-mode, and URET
copies UPIE into UIE, then sets UPIE.
\begin{commentary}
  UPIE is set after the UPIE/UIE stack is popped to enable interrupts
  and help catch coding errors.
\end{commentary}

\subsection{User Interrupt Registers ({\tt uip} and {\tt uie})}

The {\tt uip} register is a UXLEN-bit read/write register containing
information on pending interrupts, while {\tt uie} is the corresponding
UXLEN-bit read/write register containing interrupt enable bits.

\begin{figure*}[h!]
{\footnotesize
\begin{center}
\setlength{\tabcolsep}{4pt}
\begin{tabular}{KcFcFc}
\instbitrange{UXLEN-1}{9} &
\instbit{8} &
\instbitrange{7}{5} &
\instbit{4} &
\instbitrange{3}{1} &
\instbit{0} \\
\hline
\multicolumn{1}{|c|}{\wpri} &
\multicolumn{1}{c|}{UEIP} &
\multicolumn{1}{c|}{\wpri} &
\multicolumn{1}{c|}{UTIP} &
\multicolumn{1}{c|}{\wpri} &
\multicolumn{1}{c|}{USIP} \\
\hline
UXLEN-9 & 1 & 3 & 1 & 3 & 1 \\
\end{tabular}
\end{center}
}
\vspace{-0.1in}
\caption{User interrupt-pending register ({\tt uip}).}
\label{uipreg}
\end{figure*}

\begin{figure*}[h!]
{\footnotesize
\begin{center}
\setlength{\tabcolsep}{4pt}
\begin{tabular}{KcFcFc}
\instbitrange{UXLEN-1}{9} &
\instbit{8} &
\instbitrange{7}{5} &
\instbit{4} &
\instbitrange{3}{1} &
\instbit{0} \\
\hline
\multicolumn{1}{|c|}{\wpri} &
\multicolumn{1}{c|}{UEIE} &
\multicolumn{1}{c|}{\wpri} &
\multicolumn{1}{c|}{UTIE} &
\multicolumn{1}{c|}{\wpri} &
\multicolumn{1}{c|}{USIE} \\
\hline
UXLEN-9 & 1 & 3 & 1 & 3 & 1 \\
\end{tabular}
\end{center}
}
\vspace{-0.1in}
\caption{User interrupt-enable register ({\tt uie}).}
\label{uiereg}
\end{figure*}

Three types of interrupts are defined: software interrupts, timer interrupts,
and external interrupts.  A user-level software interrupt is triggered
on the current hart by writing 1 to its user software interrupt-pending
(USIP) bit in the {\tt uip} register.  A pending user-level software
interrupt can be cleared by writing 0 to the USIP bit in {\tt uip}.
User-level software interrupts are disabled when the USIE bit in the
{\tt uie} register is clear.

The ABI should provide a mechanism to send interprocessor interrupts to other
harts, which will ultimately cause the USIP bit to be set in the recipient
hart's {\tt uip} register.

All bits besides USIP in the {\tt uip} register are read-only.

A user-level timer interrupt is pending if the UTIP bit in the {\tt uip}
register is set.  User-level timer interrupts are disabled when the UTIE
bit in the {\tt uie} register is clear.  The ABI should provide a
mechanism to clear a pending timer interrupt.

A user-level external interrupt is pending if the UEIP bit in the
{\tt uip} register is set.  User-level external interrupts are disabled
when the UEIE bit in the {\tt uie} register is clear.  The ABI
should provide facilities to mask, unmask, and query the cause of external
interrupts.

The {\tt uip} and {\tt uie} registers are subsets of the {\tt mip} and {\tt
mie} registers.
Reading any field, or writing any writable field, of {\tt uip}/{\tt uie}
effects a read or write of the homonymous field of {\tt mip}/{\tt mie}.
If S-mode is implemented, the {\tt uip} and {\tt uie} registers are also
subsets of the {\tt sip} and {\tt sie} registers.

\subsection{Machine Trap Delegation Registers ({\tt medeleg} and {\tt mideleg})}

In systems with the N extension, the {\tt medeleg} and {\tt mideleg}
registers, described in Chapter~\ref{machine},  must be implemented.

In systems that implement S-mode, {\tt medeleg} and {\tt mideleg}
behave as described in Chapter~\ref{machine}.
In systems with only M and U privilege modes, setting a bit in {\tt medeleg}
or {\tt mideleg} delegates the corresponding trap in U-mode to the U-mode trap
handler.

\subsection{Supervisor Trap Delegation Registers ({\tt sedeleg} and {\tt sideleg})}

For systems with both S-mode and the N extension, new CSRs {\tt
sedeleg} and {\tt sideleg} are added.
These registers have the same layout as the machine trap delegation registers,
{\tt medeleg} and {\tt mideleg}.

{\tt sedeleg} and {\tt sideleg} allow S-mode to delegate traps to U-mode.
Only bits corresponding to traps that have been delegated to S-mode are
writable; the others are hardwired to zero.
Setting a bit in {\tt sedeleg} or {\tt sideleg} delegates the corresponding
trap in U-mode to the U-mode trap handler.

\subsection{Other CSRs}

The {\tt uscratch}, {\tt uepc}, {\tt ucause}, {\tt utvec}, and {\tt utval}
CSRs are defined analogously to the {\tt mscratch}, {\tt mepc}, {\tt mcause},
{\tt mtvec}, and {\tt mtval} CSRs.

\begin{commentary}
  A more complete writeup is to follow.
\end{commentary}

\section{N Extension Instructions}

The URET instruction is added to perform the analogous function to
MRET and SRET.

\section{Reducing Context-Swap Overhead}

The user-level interrupt-handling registers add considerable state to
the user-level context, yet will usually rarely be active in normal
use.  In particular, {\tt uepc}, {\tt ucause}, and {\tt utval} are
only valid during execution of a trap handler.

An NS field can be added to {\tt mstatus} and {\tt sstatus} following
the format of the FS and XS fields to reduce context-switch overhead
when the values are not live.  Execution of URET will place the {\tt
  uepc}, {\tt ucause}, and {\tt utval} back into initial state.
